The CrypTool Script: \\ Cryptography, Mathematics and more \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free eLearning program CrypTool\\ (8th edition -- distributed with CrypTool version 1.4.10)} \vspace{60pt}

### 2007-07-02 script/en/chapters/introduction.tex
Update scripts: chap 8, appendix a.4, quotes, minor corrections, minor layout improvements

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
Again this edition of the script amended and updated some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers
        (Mersenne and Fermat primes, ``M44'' or the biggest non-Mersenne prime
         found in March 2007 when investigating Sierpinski numbers) \\ 
	(chap. \ref{spezialzahlentypen}, \ref{RecordPrimes};
         table \ref{L_n_Largest_Known-Primes}, 
         page \pageref{L_n_Largest_Known-Primes}), 
  \item the factorisation of big numbers like ``C307''\index{C307} 
        (chap. \ref{C307}),
  \item the current record (April 2007) calculating the discrete logarithm 
        (The used sample power consisted of the first 160 digits of Euler's
        number e and the calculation was done modulo a 160 digit prime)
        (chap. \ref{L_Discrete_Logarithm}),
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}
Newly added is chapter \ref{Chapter_Crypto2020} by Prof. Buchman et al.:
{\bf Krypto2020}.

@@

At this point I'd like to thank explicitly the following six people who
particularly contributed to CrypTool\index{CrypTool}. Without their special
talents and engagement CrypTool would not be what it is today:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Prof. Dr. Claudia Eckert and  Mr.\ Thomas Buntrock.
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

>>>>

%This edition of the script adds some topics, but mainly updates areas (e.g. the
%summaries of topical research areas):
Again this edition of the script amended and updated some topics, e.g.:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers
        (Mersenne and Fermat primes, ``M44'' or the biggest non-Mersenne prime
         found in March 2007 when investigating Sierpinski numbers) \\ 
	(chap. \ref{spezialzahlentypen}, \ref{RecordPrimes};
         table \ref{L_n_Largest_Known-Primes}, 
         page \pageref{L_n_Largest_Known-Primes}), 
  \item the factorisation of big numbers like ``C307''\index{C307} 
        (chap. \ref{C307}),
  \item the current record (April 2007) calculating the discrete logarithm 
        (The used sample power consisted of the first 160 digits of Euler's
        number e and the calculation was done modulo a 160 digit prime)
        (chap. \ref{L_Discrete_Logarithm}),
%  \item progress in cryptanalysis of hash algorithms 
%        (chap. \ref{collision-attacks-against-sha-1}) and
%  \item progress in ideas for new crypto methods (RSA successor) 
%        (chap. \ref{xxxxxxxxxBrute-force-gegen-Symmetr})\index{xxxxxxxxxxxxxxxx} and
  \item the list of movies or novels, in which cryptography or number theory 
        played major role (see appendix \ref{s:appendix-movies});
        and where primes are used as hangers  
        (see curiouses in \ref{HT-Quaint-curious-Primes-usage}).
\end{itemize}
Newly added is chapter \ref{Chapter_Crypto2020} {\bf Crypto 2020}
by Prof. Buchman et al. and the appendix \ref{s:appendix-Learn-NT} 
about the {\bf learning tool for number theory}.

@@

At this point I'd like to thank explicitly the following five people who
particularly contributed to CrypTool\index{CrypTool}. Without their special
talents and engagement CrypTool would not be what it is today:
\vspace{-7pt}
%\begin{itemize}
\begin{list}{\textbullet}{\addtolength{\itemsep}{-0.5\baselineskip}}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org-Cornelius Schneider
   \item Mr.\ Florian Marchal
   \item Dr.\ Peer Wichmann
   \item Prof. Dr. Claudia Eckert and her staff (Clausius, Buntrock).
\end{list}
%\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

### 2007-07-02 script/en/chapters/moderncryptography.tex
Update scripts: chap 8, appendix a.4, quotes, minor corrections, minor layout improvements

\newpage
%********************************************************************
\hypertarget{Chapter_ModernCryptography}{}
\section{The Mathematical Ideas behind Modern Cryptography}
\label{Chapter_ModernCryptography}
%********************************************************************
(Oyono R. / Esslinger B./ Schneider J., Sep. 2000; 
            Updates Nov. 2000, Feb. 2003, Apr. 2007)

@@

\newpage
%--------------------------------------------------------------------
\subsection{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!public key}
%--------------------------------------------------------------------

@@

\fbox{\parbox{12cm}{        
\begin{tabbing}
\hspace*{0.5cm} \= \hspace*{0.5cm} \= \hspace*{0.5cm} \= \kill
\>{\bf for} $ i=n $ {\bf to} 1 {\bf do} \\
\>\> {\bf if} $ T\geq a_i $ {\bf then}\\
\>\> \> $ T:=T-s_i $ \\
\>\>\> $ x_i:=1 $ \\
\>\> {\bf else} \\
\>\>\> $ x_i:=0 $ \\
\>{\bf if} $ T=0 $ {\bf then} \\
\>\> $ X:=(x_1, \dots, x_n) $ is the solution. \\
\>{\bf else} \\
\>\> No solution exists.
\end{tabbing}
}}
\end{center}
\vskip +10 pt
{\bf Algorithm 1.} Solving knapsack problems with super-increasing weights
\vskip +20 pt

%--------------------------------------------------------------------
\subsubsection{Merkle-Hellman knapsack encryption}
%--------------------------------------------------------------------
\index{Hellman, Martin} \index{Merkle, Ralph}

@@

\begin{center}
\fbox{\parbox{12cm}{        
Let $ (a_1, \dots, a_n) $ be super-increasing. Let $ m $ and $ w $ be two co-prime 
numbers with $ m > \sum_{i=1}^{n} a_i $ and $ 1\leq w \leq m-1. $ Select 
$\bar{w} $ with $ w \bar{w} \equiv 1 \mod m $ the modular inverse of $ w $ and 
set $ b_i:= wa_i \mod m, $ $ 0\leq b_i < m $ for $ i=1, \dots ,n, $ and verify 
whether the sequence $ b_1, \dots b_n $ is not super-increasing. A permutation $ 
b_{\pi (1)}, \dots , b_{\pi(n)} $ of  $ b_1, \dots , b_n $ is then published and 
the inverse permutation $ \mu $ to $ \pi $ is defined secretly. A sender writes 
his/her message in blocks $ (x_1^{(j)}, \dots, x_n^{(j)}) $ of binary numbers $ 
n $ in length, calculates
\[ g^{(j)}:= \sum_{i=1}^n x_{i}^{(j)} b_{\pi(i)} \]
and sends $ g^{(j)}, (j=1,2, \dots). $\par
The owner of the key calculates
\[ G^{(j)}:=\bar{w} g^{(j)} \mod m ,\quad 0 \leq G^{(j)} < m \]
and obtains the $ x_{\mu(i)}^{(j)} \in \{ 0,1\} $ (and thus also the $ x_i^{(j)} 
$) from
\begin{eqnarray*}
G^{(j)} & \equiv & \bar{w} g^{(j)} = \sum_{i=1}^n x_i^{(j)} b_{\pi (i)} \bar{w} 
\equiv \sum_{i=1}^n x_i^{(j)} a_{\pi (i)} \mod m \\
& = & \sum_{i=1}^n x_{\mu (i)}^{(j)} a_{\pi (\mu (i))} = \sum _{i=1}^n x_{\mu 
(i)}^{(j)} a_i \mod m, 
\end{eqnarray*}
by solving the easier 0-1 knapsack problems $ K(a_1,\dots,a_n;G^{(j)}) $ with 
super-increasing sequence $ a_1, \dots,a_n $.
}}
\end{center}
\vskip +10 pt
{\bf Merkle-Hellman procedure} (based on knapsack problems).
\vskip +20 pt

@@

\vskip +10 pt
{\bf RSA procedure} (based on the factorisation problem).
\vskip +20 pt
\index{Factorisation!factorisation problem}
\index{Euler!(phi) function}
{\bf Comment:} 
The Euler phi function is defined as: $ \phi (N)$ is the number of natural 
numbers that do not have a common factor with $ N $ \index{Co-prime} {}
$ x \leq N. $ Two natural numbers $ a $ and $ b $ are co-prime if 
$ \ggt (a,b)=1. $

@@

\vskip +10 pt
{\bf Rabin procedure} (based on the factorisation problem).
\vskip +20 pt

@@

%--------------------------------------------------------------------
\subsection{The discrete logarithm as a basis for public key procedures}
Discrete logarithms \index{Discrete logarithm} form the basis for a large number of algorithms for public-
key procedures.
\index{Logarithm problem!discrete}

@@

Let $ p $ be a prime number and let $ g \in \Z_p^\ast=\{0,1,\ldots,p-1\} $. Then 
the discrete exponential function base $ g $  is defined as
\[ e_g : k \longrightarrow y:=g^k \mod p, \quad 1\leq k \leq p-1. \]
The inverse function is called a discrete logarithm function $ \log_g $; the 
following holds:
\[ \log_g (g^k) =k. \]
\index{Exponential function!discrete} The problem of the discrete logarithm (in 
$ \Z_p^\ast$) is understood to be as follows:
\[ {\rm Given~ } p,g {\rm ~and~ } y, {\rm~determine~} k {\rm ~such~that~ } 
y=g^k \mod p {\rm .}\]
It is much more difficult to calculate the discrete logarithm than to evaluate 
the discrete exponential function (see chapter \ref{MultOrdPrimitveRoot}).
There are several procedures for calculating the discrete logarithm 
\cite{4Stinson1995}\index{Stinson 1995}:
\vskip + 5pt
\begin{center}
\begin{tabular}{|l|l|}\hline
Name                 &        Complexity \\ \hline \hline
Baby-Step-Giant-Step &         $ O(\sqrt{p}) $ \\ \hline
Silver-Pohlig-Hellman    &    polynomial in $ q, $ the greatest \\
&  prime factor of $ p-1. $ \\ \hline
Index-Calculus &             $ O(e^{(1+o(1)) \sqrt{\ln (p) \ln (\ln (p))}}) $ \\ 
\hline
\end{tabular}
\end{center}
\vskip +5 pt
\index{Silver} \index{Pohlig, S. C.} \index{Hellman, Martin}
\index{Baby-Step-Giant-Step}

@@

\vskip +10 pt
{\bf Diffie-Hellman key agreement.}
\vskip +20 pt

@@

\vskip +10 pt
{\bf ElGamal procedure} (based on the factorisation problem).
\vskip +20 pt

@@

\vskip +10 pt
{\bf Generalised ElGamal procedure} (based on the factorisation problem).
\vskip +20 pt

>>>>

\newpage
%********************************************************************
\hypertarget{Chapter_ModernCryptography}{}
\section[The Mathematical Ideas behind Modern Cryptograph]
        {The Mathematical Ideas behind Modern Cryptograph\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can apply
    some of the methods introduced here (RSA, Rabin, DH, ElGamal)
    (see learning unit 4.2 and 4.3, pages 9-17/17).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{Chapter_ModernCryptography}
%********************************************************************
(Oyono R. / Esslinger B./ Schneider J., Sep. 2000; 
            Updates Nov. 2000, Feb. 2003, Apr. 2007)

\vskip +30 pt

\begin{center}
\fbox{\parbox{15cm}{
   {\em Georg Christoph Lichtenberg\footnotemark:}\\
   I don't know if its getting better, if we change it,\\
   but I know, that we have to change it, if it should become better.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  Georg Christoph Lichtenberg\index{Lichtenberg, Georg Christoph}, 
  German writer and physicist (1742-1799),\\ 
  (also see: \href{http://en.wikipedia.org/wiki/Georg\_Christoph\_Lichtenberg}
              {\tt http://en.wikipedia.org/wiki/Georg\_Christoph\_Lichtenberg})
}

@@

%\newpage
\vskip +20 pt
%--------------------------------------------------------------------
\subsection{Knapsack problem as a basis for public key procedures}%
\index{Cryptography!public key}
%--------------------------------------------------------------------

@@

\fbox{\parbox{12cm}{        
\begin{tabbing}
\hspace*{0.5cm} \= \hspace*{0.5cm} \= \hspace*{0.5cm} \= \kill
\>{\bf for} $ i=n $ {\bf to} 1 {\bf do} \\
\>\> {\bf if} $ T\geq a_i $ {\bf then}\\
\>\> \> $ T:=T-s_i $ \\
\>\>\> $ x_i:=1 $ \\
\>\> {\bf else} \\
\>\>\> $ x_i:=0 $ \\
\>{\bf if} $ T=0 $ {\bf then} \\
\>\> $ X:=(x_1, \dots, x_n) $ is the solution. \\
\>{\bf else} \\
\>\> No solution exists.
\end{tabbing}
}}
\end{center}
%\vskip +10 pt
{\bf Algorithm 1.} Solving knapsack problems with super-increasing weights

\vskip +40 pt
%--------------------------------------------------------------------
\subsubsection{Merkle-Hellman knapsack encryption}
%--------------------------------------------------------------------
\index{Hellman, Martin} \index{Merkle, Ralph}

@@

\begin{center}
\fbox{\parbox{12cm}{        
Let $ (a_1, \dots, a_n) $ be super-increasing. Let $ m $ and $ w $ be two co-prime 
numbers with $ m > \sum_{i=1}^{n} a_i $ and $ 1\leq w \leq m-1. $ Select 
$\bar{w} $ with $ w \bar{w} \equiv 1 \mod m $ the modular inverse of $ w $ and 
set $ b_i:= wa_i \mod m, $ $ 0\leq b_i < m $ for $ i=1, \dots ,n, $ and verify 
whether the sequence $ b_1, \dots b_n $ is not super-increasing. A permutation $ 
b_{\pi (1)}, \dots , b_{\pi(n)} $ of  $ b_1, \dots , b_n $ is then published and 
the inverse permutation $ \mu $ to $ \pi $ is defined secretly. A sender writes 
his/her message in blocks $ (x_1^{(j)}, \dots, x_n^{(j)}) $ of binary numbers $ 
n $ in length, calculates
\[ g^{(j)}:= \sum_{i=1}^n x_{i}^{(j)} b_{\pi(i)} \]
and sends $ g^{(j)}, (j=1,2, \dots). $\par
The owner of the key calculates
\[ G^{(j)}:=\bar{w} g^{(j)} \mod m ,\quad 0 \leq G^{(j)} < m \]
and obtains the $ x_{\mu(i)}^{(j)} \in \{ 0,1\} $ (and thus also the $ x_i^{(j)} 
$) from
\begin{eqnarray*}
G^{(j)} & \equiv & \bar{w} g^{(j)} = \sum_{i=1}^n x_i^{(j)} b_{\pi (i)} \bar{w} 
\equiv \sum_{i=1}^n x_i^{(j)} a_{\pi (i)} \mod m \\
& = & \sum_{i=1}^n x_{\mu (i)}^{(j)} a_{\pi (\mu (i))} = \sum _{i=1}^n x_{\mu 
(i)}^{(j)} a_i \mod m, 
\end{eqnarray*}
by solving the easier 0-1 knapsack problems $ K(a_1,\dots,a_n;G^{(j)}) $ with 
super-increasing sequence $ a_1, \dots,a_n $.
}}
\end{center}
%\vskip +10 pt
{\bf Merkle-Hellman procedure} (based on knapsack problems).
\vskip +20 pt

@@

%\vskip +10 pt
{\bf RSA procedure} (based on the factorisation problem).
\vskip +20 pt
\index{Factorisation!factorisation problem}
\index{Euler!(phi) function}
{\bf Comment:} 
The Euler phi function is defined as: $ \phi (N)$ is the number of natural 
numbers that do not have a common factor with $ N $ \index{Co-prime} {}
$ x \leq N. $ Two natural numbers $ a $ and $ b $ are co-prime if 
$ \ggt (a,b)=1. $

@@

%\vskip +10 pt
{\bf Rabin procedure} (based on the factorisation problem).
\vskip +20 pt

@@

\vskip +20 pt
%--------------------------------------------------------------------
\hypertarget{HT-Discrete-Logarithm-as-Basis}{}
\subsection[The discrete logarithm as basis for public key procedures]
           {The discrete logarithm as basis for public key procedures\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can play with the
    distribution of the discrete logarithm values and apply Shank's
    baby-step-giant-step method:
    See learning units 6.1-6.3, pages 1-6/6.\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix \ref{s:appendix-Learn-NT}.
}
\label{L-Discrete-Logarithm-as-Basis}
Discrete logarithms \index{Discrete logarithm} form the basis for a large
number of algorithms for public-key procedures.
\index{Logarithm problem!discrete}

@@

Let $ p $ be a prime number and let $ g \in \Z_p^\ast=\{0,1,\ldots,p-1\} $. Then 
the discrete exponential function base $ g $  is defined as
\[ e_g : k \longrightarrow y:=g^k \mod p, \quad 1\leq k \leq p-1. \]
The inverse function is called a discrete logarithm function $ \log_g $; the 
following holds:
\[ \log_g (g^k) =k. \]
\index{Exponential function!discrete} The problem of the discrete logarithm (in 
$ \Z_p^\ast$) is understood to be as follows:
\[ {\rm Given~ } p,g {\rm ~and~ } y, {\rm~determine~} k {\rm ~such~that~ } 
y=g^k \mod p {\rm .}\]
It is much more difficult to calculate the discrete logarithm than to evaluate 
the discrete exponential function (see chapter \ref{MultOrdPrimitveRoot}).
There are several procedures for calculating the discrete logarithm 
\cite{4Stinson1995}\index{Stinson 1995}:
\vskip + 5pt
\begin{center}
\begin{tabular}{|l|l|}\hline
Name                 &        Complexity \\ \hline \hline
Babystep-Giantstep &         $ O(\sqrt{p}) $ \\ \hline
Silver-Pohlig-Hellman    &    polynomial in $ q, $ the greatest \\
&  prime factor of $ p-1. $ \\ \hline
Index-Calculus &             $ O(e^{(1+o(1)) \sqrt{\ln (p) \ln (\ln (p))}}) $ \\ 
\hline
\end{tabular}
\end{center}
\vskip +5 pt
\index{Silver} \index{Pohlig, S. C.} \index{Hellman, Martin}
\index{Baby-Step-Giant-Step}

@@

%\vskip +10 pt
{\bf Diffie-Hellman key agreement.}
\vskip +20 pt

@@

%\vskip +10 pt
{\bf ElGamal procedure} (based on the factorisation problem).
\vskip +20 pt

@@

%\vskip +10 pt
{\bf Generalised ElGamal procedure} (based on the factorisation problem).
\vskip +20 pt
